---
title: Lazily materializing assay rows or columns
author:
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
date: "Revised: February 22, 2019"
output:
  BiocStyle::html_document
package: LazyAssayVectors
vignette: >
  %\VignetteIndexEntry{Lazy assay vectors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

# Motivation

The `r Biocpkg("LazyAssayVectors")` package allows users and package developers to represent an assay matrix as a `data.frame` of lazily materialized vectors.
This is occasionally useful for efficient integration with ecosystems that expect `data.frame`s, the most obvious being `r CRANpkg("ggplot2")`.
Consider this very large matrix:

```{r}
library(Matrix)
y <- rsparsematrix(1e6, 1e6, density=0.0000001)
rownames(y) <- sprintf("Gene_%i", seq_len(nrow(y)))
```

Let's say that we want to represent it as a `data.frame` of column vectors.
The theoretical size of such a `data.fram` easily exceeds the available memory on most machines,
yet here we are:

```{r}
library(LazyAssayVectors)
df <- createLazyRows(y)
object.size(df)
```

And why would we want to represent it as a `data.frame`?
Well, it allows us to do stuff like the following:

```{r, fig.show="hide"}
library(ggplot2)
ggplot(df) + geom_point(aes(x=Gene_2, y=Gene_1))
```

The key aspect of this is that _everything_ from `y` is available in the `df` and can be used by downstream methods -
in this case, the various `geom`s and `stat`s.
We do not have to extract out `Gene_1` and `Gene_2` from `y` to put in `df` and then specify `Gene_1` and `Gene_2` **again** in the layers; rather, a single specification is sufficient, which makes it easier to rapidly write `r Biocpkg("ggplot")` code.
This applies to any function that expects a `data.frame` as input:

```{r}
design <- model.matrix(~Gene_10000, data=df)
dim(design)
```

# Supported matrix representations

Every matrix representation is supported provided it yields numeric, integer or logical values upon `[`.
Let's demonstrate with some file-backed matrices from the `r Biocpkg("HDF5Array")` package.

```{r}
library(TENxBrainData)
sce <- TENxBrainData()
dim(sce)
```

In this case, we'll work with the counts and create a column-wise `data.frame`.
Again, we can trick R into loading an object that is much bigger than the available RAM (at least on my machine).

```{r}
df <- createLazyColumns(counts(sce))
object.size(df)
```

We are able to extract columns in the `data.frame`, which triggers lazy materialization from the underlying `HDF5Matrix`.
It must be said, though, that this is a fair bit slower than our previous example due to the delay in reading from file.

```{r}
summary(df[,1])
```

On rare occasions, you may see a timeout warning that will instruct you to run `restartWatcher()`.
This will restart the watcher process that actually handles data reading from file.

# Caveats on laziness

Lazily materialized vectors are powerful as we can effectively extend any vector-ingesting function to work on large datasets.
However, as Superman once said, with great power comes great responsibility.
In particular, users of lazy vectors have to take care to avoid inadvertently materializing the entirety of the `data.frame`,
which is surprisingly common with operations like subsetting.
This often requires some protection to be placed around the lazy vectors to avoid unintended materialization by end-users;
for example, the `ggcells()` function from `r Biocpkg("scater")` makes it harder to directly manipulate the `data.frame` by wrapping it in a `ggplot()` call.

<!--
If you wanted to file a PR about it being Uncle Ben who said it, you've already
fallen into my trap. Defeat, like coffee, is best served bitter.
-->

Another materialization trigger occurs when lazy vectors are saved,
either by explicitly by `saveRDS()` or implicitly by `save()` (and higher-level functions like `r CRANpkg("knitr")`'s caching).
To avoid this, we provide the `dropLazyColumns()` function to remove all non-materialized lazy assay vectors from a `data.frame`.
This allows us to conveniently operate on the `data.frame` as usual and,
once we are finished, we can simply drop all unused columns to obtain a slimmer object for saving.
Again, it is unlikely that end-users will call `dropLazyColumns()` directly;
rather, it will be called by other functions to ensure that any user-exposed objects are cut down to size.

# Session information

```{r}
sessionInfo()
```
